\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-07 doc/incertae_sedis.tex
minor

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
    \item An \emph{incertae sedis} node may intrude into its siblings and their descendants without conflict.
    \item An \emph{incertae sedis} node has an unrestricted range within the descendants of its sibling taxa.
    \item The semantics of an \emph{incertae sedis} node does not depend on side-information such as taxon ranks, but only on the taxonomy tree.
    \item Semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
      % These could be goals, but maybe discuss them as side-effects:
      % Side effect #1. Don't interdigitate
      % Side effect #2. Allow placing A next to B, and then placing B within A. (See Fig. 6)
\end{enumerate}

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.

\subsubsection{Unrestricted range and ignoring side information}
The choice to allow each \emph{incertae sedis} taxon to attach anywhere within
the descendants of its siblings requires some explanation, for two reasons.
First, more precise information about where a taxon can attach is easy to
incorporate. If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
set for each $p\in\mathcal{P}$.  Such an approach would require that the path from $x$ to $p$
go through only other potential ancestors.

@@

Unfortunately, reliable information about the range of valid attachment points
for each \emph{incertae sedis} taxon is lacking.  Ideally, the \emph{incertae
  sedis} annotation for some taxon $x$ would be accompanied by a specific statement
about which other taxa might include $x$.  However, OTT does not provide this
directly.  Constructing a range for each incertae sedis taxon relies on detailed
readings of character argumentation (which is lacking in OTT) or rank-based
information (which is frequently untrustworthy in OTT). Thus, here we pursue
an approach of ascribing a meaning to the \emph{incertae sedis} label which
attempts to capture the core of the idea that it articulates, but which can be
applied automatically across the taxonomy without additional information.

\subsubsection{Terminology for rooted splits}
\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}

\paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each edge by noting that
removing the edge would produce two trees. The leaf label sets of these trees
the partition the full leaf set $\mathcal{L}$ into two groups. The include set $\mathcal{I}(e)$ which
does not contain the root, and the exclude set $\mathcal{E}(e)$ which does contain the root.
Such a split may be written
\[ \mathcal{I}(e)|\bullet\mathcal{E}(e).\]
where the $\bullet$ indicates the root. The include set is the set of leaves that are
descendants of the daughter node of the edge. The exclude set is the set of
leaves that are not descendants of that node. If no taxa are \emph{incertae sedis},
then the exclude set for a node is just the total tip set minus the include set
for the node: $\mathcal{E}(e)  =\mathcal{L}-\mathcal{I}(e)$. For a node $n$ on the tip-ward side of an edge $e$, we
may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$. For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
then the leaf labels that are descendants of $n$ will be in the include set for
$n$ and all its ancestors and will be in the exclude set for all other nodes/edges.

In terms of phylogenetic nomenclature the supertree pipeline acts as if the
taxonomy defines a branch-based name \citep[see][]{deQueiroz2013} for each
higher taxon of the form. So for the name labeling node $n$ in the taxonomic
tree, the operational definition would be: ``the clade defined by all taxa that
are more closely related to $\mathcal{I}(n)$ than to $\mathcal{E}(n)$, if such a clade exists on the tree.''

@@

An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
moved into the descendants of any of its siblings.  Thus, annotating $n$ as
\emph{incertae sedis} really modifies the meanings of its siblings so that they
no longer exclude $n$.  In terms of rooted splits, we can represent this lack of
exclusion by omitting taxa from the exclude set for a node.  Such a split does
not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.
Thus, when interpreted correctly the \emph{incertae sedis} 
annotations are reducing the amount of information that the taxonomy brings to
the supertree problem.

We seek to define modified splits for each branch of the taxonomy tree. The
include sets of these splits remain unchanged, but the exclude sets may be reduced
\emph{incertae sedis} taxa to intrude into their sibling taxa.  Therefore, a
semantics can be given simply in terms of the exclude set for each taxonomy node.

>>>>

We seek a semantics for supertrees with \emph{incertae sedis }taxa
that satisfy the following properties:
\begin{enumerate}
    \item An \emph{incertae sedis} node may intrude into its siblings and their descendants without conflict.
    \item An \emph{incertae sedis} node has an unrestricted range within the descendants of its sibling taxa.
    \item The semantics of an \emph{incertae sedis} node does not depend on additional information such as taxon ranks, but only on the taxonomy tree.
    \item The semantics is based on deriving a rooted (partial) split for each branch of the taxonomy.
      % These could be goals, but maybe discuss them as side-effects:
      % Side effect #1. Don't interdigitate
      % Side effect #2. Allow placing A next to B, and then placing B within A. (See Fig. 6)
\end{enumerate}

%\subsubsection{Terminology for rooted splits}

The rooted split associated with taxonomy node $n$ serves to determine both when $n$ is in conflict
with an input tree, and to where to place the taxon name for clade $n$ on the synthesis tree.
% Removed mention of incertae sedis possibly affecting subproblem decomposition.  It
% actually doesn't but that is surprising, and therefore kind of a result.
% \paragraph{Definition of rooted splits}
As is typically done, we define rooted splits for each node by noting that
removing the edge connecting the node to its parent would produce two trees.
The leaf label sets of these trees
the partition the full leaf set $\mathcal{L}$ into two groups.
The include set $\mathcal{I}(n)$ which
does not contain the root, and the exclude set $\mathcal{E}(n)$ which does contain the root.
Such a split may be written
\[ \mathcal{I}(n)|\bullet\mathcal{E}(n).\]
where the $\bullet$ indicates the root. 
The include set is the set of labels for leaves that are
descendants of the node.
The exclude set is the set of
labels for leaves that are not descendants of that node. 
If no taxa are \emph{incertae sedis},
then the exclude set for a node is just the total tip set minus the include set
for the node: $\mathcal{E}(n)  =\mathcal{L}-\mathcal{I}(n)$.
%For a node $n$ on the tip-ward side of an edge $e$, we
%may also write $\mathcal{I}(n)$ for $\mathcal{I}(e)$, and $\mathcal{E}(n)$ for $\mathcal{E}(e)$.
For any two rooted splits $A=A_{1}|\bullet A_{2}$
and $B=B_{1}|\bullet B_{2}$ , we say that $A$ \emph{displays }$B$ if $B_{1}\subseteq A_{1}$ and $B_{2}\subseteq A_{2}$.

Note that when a taxon rooted at node $n$ is not marked as \emph{incertae sedis},
then the leaf labels that are descendants of $n$ will be in the include set for
$n$ and all its ancestors and will be in the exclude set for all other nodes.

In terms of phylogenetic nomenclature the supertree pipeline acts as if the
taxonomy defines a conditional node-based name \citep[see][]{deQueiroz2013} for each
higher taxon.
So for the name labeling node $n$ in the taxonomic
tree, the operational definition would be: ``the clade defined by the most recent
common ancestor (MRCA) of $\mathcal{I}(n)$ as long as that MRCA is not an ancestor of
$\mathcal{E}(n)$.''
OTT is not based on a set of explicitly defined phylogenetic name definitions.
In many contexts, a branch-based definition for $n$ (``the clade by all of the taxa more
closely related to all fo them members of $\mathcal{I}(n)$ than to any member of
$\mathcal{E}(n)$, if such a clade exists'') would result in similar behavior
during the tree construction. 
An exception to this (discussed below) arises in the step of assigning names to the 
    supertree.

\paragraph{Equivalence of a taxonomy tree and a list of rooted splits}

\subsubsection{Unrestricted range and ignoring additional information}
The choice to allow each \emph{incertae sedis} taxon to attach anywhere within
the descendants of its siblings requires some explanation, for two reasons.
First, more precise information about where a taxon can attach is easy to
incorporate. 
If $\mathcal{P}$ is the set of taxa that are potential ancestors for \emph{incertae
sedis} taxon $x$, then we may simply remove the descendants of $x$ from the exclude
set for each $p\in\mathcal{P}$.  Such an approach would require that the path from $x$ to $p$
go through only other potential ancestors.

@@

Unfortunately, reliable information about the range of valid attachment points
for each \emph{incertae sedis} taxon is lacking in OTT and most other large-scale
taxonomies.
Ideally, the \emph{incertae
  sedis} annotation for some taxon $x$ would be accompanied by a specific statement
about which other taxa might include $x$.  However, OTT does not provide this
directly.  Constructing a range for each incertae sedis taxon relies on detailed
readings of character argumentation (which is lacking in OTT) or rank-based
information (which is frequently untrustworthy in OTT). Thus, here we pursue
an approach of ascribing a meaning to the \emph{incertae sedis} label which
attempts to capture the core of the idea that it articulates, but which can be
applied automatically across the taxonomy without additional information.

@@

An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
moved into the descendants of any of its siblings.
Thus, annotating $n$ as
\emph{incertae sedis} really modifies the meanings of the taxanomic statements
that of the siblings of $n$ and all taxonomic nodes that descend from those siblings.
 so that they
no longer exclude $n$.
If we denote this set of taxa (the set of taxa that are {\emph{potential}} 
ancestors for \emph{incertae sedis} taxon $n$) as $\mathcal{P}_n$, 
then we may simply remove the descendants of $n$ from the exclude
set for each $p\in\mathcal{P}_n$.
%In terms of rooted splits, we can represent this lack of
%   exclusion by omitting taxa from the exclude set for a node.
% We seek to define modified splits for each branch of the taxonomy tree.
The include sets of these splits remain unchanged, but the exclude sets may be reduced
\emph{incertae sedis} taxa to intrude into their sibling taxa.
% Therefore, a semantics can be given simply in terms of the exclude set for each taxonomy node.
Such a split does
not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.
Thus, when interpreted correctly the \emph{incertae sedis} 
annotations are reducing the amount of information that the taxonomy brings to
the supertree problem.
